---
title: "Weather Data Analysis"
author: "Jasper Slingsby"
date: "9 March 2017"
output: pdf_document
---

<br>

## Trends in mean monthly maximum temperature anomalies:

<br>

```{r, echo=FALSE, message=FALSE}
###Set Working directory
datwd <- ""

###Get libraries
library(ggplot2)
library(dplyr)
library(climdex.pcic)
library(MCMCglmm)
library(gdata)

###Get weather data
if(file.exists("weather.csv")) {
ddat <- read.csv(paste0(datwd,"weather.csv"), stringsAsFactors = F)
} else {
ddat <- read.xls(paste0(datwd,"Dataset S1.xlsx"), sheet = "weather", stringsAsFactors = F)
}
           
ddat$Date <- as.Date(ddat$Date, "%d/%m/%Y")

###Function to summarise data and plot
anomalyplot <- function(p, q, tit, yl) # p is a dataframe with columns "Year", "Month" and other variables, q is the variable of interest for this plot, tit is the title, and y is the y-axis label
{
###Summarise data for plotting 
x <- tbl_df(na.omit(p))
y <- summarise(group_by(x, Year, Month), MMT=mean(q)) # monthly means by year
x <- summarise(group_by(x, Month), MMMT=mean(q)) # monthly means across all years
y <- merge(y, x)

y$Positive <- y$MMT - y$MMMT # calculate anomalies
y$Positive[which(y$Positive<0)] <- 0
y$Negative <- y$MMT - y$MMMT
y$Negative[which(y$Negative>0)] <- 0

###Plot
g <- ggplot(y, aes(x = Year))
g <- g + geom_ribbon(aes(ymin = 0, ymax = Positive, fill="Positive")) + geom_ribbon(aes(ymin = Negative, ymax = 0 ,fill="Negative"))
g <- g + scale_fill_manual(name="",values=c("#4393C3","#D6604D"))
g <- g + facet_grid(.~Month)
g + ggtitle(tit) + coord_flip() + ylab(yl)
}

anomalyplot(p=ddat[,c("Year","Month","Temp")], q=alist(Temp)[[1]], tit="Mean monthly Max Temp Anomaly (1961-2010)", yl="Degrees Celcius")

```

<br>

Pretty clear increases in positive anomalies.

<br>

\pagebreak

<br>

## Trends in cumulative monthly rainfall anomalies:

<br>

```{r, echo=FALSE}

###Summarise data for plotting 
x <- tbl_df(na.omit(ddat))
y <- summarise(group_by(x, Year, Month), MMP=sum(Rain))
x <- summarise(group_by(y, Month), MMMP=mean(MMP))
y <- merge(y, x)

y$Positive <- y$MMP - y$MMMP
y$Positive[which(y$Positive<0)] <- 0
y$Negative <- y$MMP - y$MMMP
y$Negative[which(y$Negative>0)] <- 0

###Plot
g <- ggplot(y, aes(x = Year))
g <- g + geom_ribbon(aes(ymin = 0, ymax = Positive, fill="Positive")) + geom_ribbon(aes(ymin = Negative, ymax = 0 ,fill="Negative"))
g <- g + scale_fill_manual(name="",values=c("#D6604D", "#4393C3"))
g <- g + facet_grid(.~Month)
g + ggtitle("Cumulative Monthly Rainfall Anomaly (1961-2010)")+coord_flip()+ylab("Rainfall (mm)")

```

<br>

Not much trend. There's bigger variance in April to September because this spans the wet season.

<br>

\pagebreak

<br>

##Analysis of maximum count of consecutive hot and dry days (1963-2009)

<br>

Let's look at the plot of consecutive hot and dry days and then model them as a function of year with MCMCglmm (default priors and normal errors).

```{r, echo=FALSE}

###Trim to April 1963 onwards
ddat <- ddat[ddat$Date>as.Date("1963-3-31"),]

###Set Dry vs Wet season
ddat$Season <- NA
ddat$Season[which(months(ddat$Date)%in%c("October","November","December","January","February","March"))] <- "Dry"
ddat$Season[which(months(ddat$Date)%in%c("April","May","June","July","August","September"))] <- "Wet"

###Set summer years" (i.e. the summer Oct 1963 to March 1964 is the summer of 1963...)
ddat$DryYear <- NA
ddat$DryYear[1] <- ddat$Year[1]
for (i in 2:nrow(ddat)){
  if(ddat$Season[i]=="Dry" & ddat$Season[i-1]=="Wet")
   ddat$DryYear[i] <- ddat$DryYear[i-1]+1
  else
    ddat$DryYear[i] <- ddat$DryYear[i-1] 
}

ddat$SeasonYear <- paste0(ddat$Season, ddat$DryYear, sep="")

###Days with Temperature > 21 and no rain.
ddat$xE <- (ddat$Temp > 21)*(ddat$Rain < 1)

###Consectutive days with Temperature > 21 and no rain
CHDseason <- spell.length.max(ddat$xE, as.factor(ddat$SeasonYear), 0, ">", spells.can.span.years=FALSE)
  
#CHDspan <- spell.length.max(ddat$xE, as.factor(ddat$Year), 0, ">", spells.can.span.years=TRUE)

###Consecutuve dry days
cdd <- spell.length.max(ddat$Rain, as.factor(ddat$SeasonYear), 1, "<", spells.can.span.years=FALSE)

###Bind with years
climstress <- data.frame(Consecutive_Hot_and_Dry_Days = CHDseason[1:46], Consecutive_Dry_Days = cdd[1:46], Year = as.integer(1963:2008))

###Plot
g <- ggplot(data=climstress, na.rm=T, aes(x=Year, y=Consecutive_Hot_and_Dry_Days)) + 
  geom_point() + 
  geom_smooth(alpha=.2, size=1, method="lm") + 
  ylab("Maximum consecutive \n hot and dry days") +
  #theme_bw() +
  #theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank()) +
  xlim(1962, 2011)

g

#ggsave(paste0(datwd,"Consecutive_Hot_and_Dry_Days", Sys.Date(), ".pdf"), width=11.4, height=6, units = "cm", dpi=600, plot=g)

summary(MCMCglmm(Consecutive_Hot_and_Dry_Days ~ Year, data=climstress, verbose = F))
```

<br>

Lastly, hidden code calculates the most extreme post fire weather (CDD and CHD) experienced in the first year after fire and outputs this as *postfireweather.csv*. This can be altered to change thresholds etc for downstream analyses if desired.

<br>

```{r, echo = F}

###Code to calculate when our sites burnt relative to extreme summer weather events

# if(file.exists("fires.csv")) {
# burns <- read.csv(paste0(datwd,"fires.csv"), row.names=1)
# } else {
# burns <- read.xls(paste0(datwd,"Dataset S1.xlsx"), sheet = "fires", row.names=1)
# }
# 
# climstress$SeasonYear <- rownames(climstress)
# 
# # Extract for each inter-survey period across full period (1966-2010)
# x <- merge(burns, climstress, by = "SeasonYear")
# y <- aggregate(x[,5:6], by=list(x$plot), FUN="max")
# colnames(y)[1] <- "Plot"
# 
# # Extract for each inter-survey period across first period (1966 - 1996)
# x66_96 <- x[-which(x$year>1995),] 
# y66_96 <- aggregate(x66_96[,5:6], by=list(x66_96$plot), FUN="max")
# colnames(y66_96) <- paste0(colnames(y66_96), "_66_96")
# colnames(y66_96)[1] <- "Plot"
# 
# # Bin into classes
# postfireweather <- merge(y66_96, y)
# 
# postfireweather$cdd66_96 <- cut(y66_96$Consecutive_Dry_Days_66_96, breaks=c(0,40,100)) #c(0,50,200)
# postfireweather$chd66_96 <- cut(y66_96$Consecutive_Hot_and_Dry_Days_66_96, breaks=c(0,5,11)) #c(0,5,9)
# 
# postfireweather$cdd66_10 <- cut(y$Consecutive_Dry_Days, breaks=c(0,40,100))
# postfireweather$chd66_10 <- cut(y$Consecutive_Hot_and_Dry_Days, breaks=c(0,5,11))
# 
# write.csv(postfireweather, paste0(datwd,"postfireweather.csv"))

```

<br>


